---
title: "Health Outcomes County Level USA"
author: "Aryaman Bedi"
date: "2023-06-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading the required libraries:

```{r}
library(arm)
library(dplyr)
library(readxl)
library(ggplot2)
library(sf)
library(mapview)
library(raster)
```

Reading the required data for finding analysing the relationship between premature age- adjusted mortality rate of a county with smoking, obesity, physical inactivity, diabetes and poor mental health

```{r}
data <- read_excel("/Users/bedi/Downloads/DATA5207_Project/Data/Q3 What predicts better health outcomes at the county level in the United States/2018 County Health Rankings Data - v2.xls", sheet = 6)


dat <- read_excel("/Users/bedi/Downloads/DATA5207_Project/Data/Q3 What predicts better health outcomes at the county level in the United States/2018 County Health Rankings Data - v2.xls", sheet = 4)
```

Selecting the required columns (i.e., the five factors mentioned above)

```{r}
d6 <- data %>%
  dplyr::select(...1,...2,...3,`Premature age-adjusted mortality`,Demographics)

d4 <- dat %>%
  dplyr::select(...1,'Poor mental health days','Adult smoking','Adult obesity','Physical inactivity','Diabetes monitoring')
```

Renaming Columns 

```{r}
d6 <- d6 %>%
  rename(FIPS = ...1,
         State = ...2,
         County = ...3,
         Population = Demographics,
         'Premature age-adjusted mortality (Deaths)' = 'Premature age-adjusted mortality')

d4 <- d4 %>%
  rename(FIPS = ...1,
         'Adult smoking (%)' = 'Adult smoking',
         'Adult obesity (%)' = 'Adult obesity',
         'Diabetics' = 'Diabetes monitoring',
         'Physical inactivity (%)' = 'Physical inactivity')
```

Removing first row from both to avoid repetion of column names in first row before joining both the datasets 

```{r}
d4 <- d4[-1, ]
d6 <- d6[-1, ]
```

Joining both the columns using the FIPS column followed by converting all the columns to numeric datatype 

```{r}
joined_data <- left_join(d6, d4, by = "FIPS")

joined_data <- joined_data %>%
  mutate(`Adult smoking (%)` = as.numeric(`Adult smoking (%)`),
         `Premature age-adjusted mortality (Deaths)` = as.numeric(`Premature age-adjusted mortality (Deaths)`),
         `Adult obesity (%)` = as.numeric(`Adult obesity (%)`),
         `Physical inactivity (%)` = as.numeric(`Physical inactivity (%)`),
         Diabetics = as.numeric(Diabetics),
         `Poor mental health days` = as.numeric(`Poor mental health days`),
         `Population` = as.numeric(`Population`))

```


Converting Premature Age Adjusted Mortality to Percentage for all the counties
```{r}
joined_data$`Premature age-adjusted mortality (Deaths)` <- ((joined_data$`Premature age-adjusted mortality (Deaths)` / joined_data$Population)) * 100
joined_data$Diabetics <- ((joined_data$Diabetics / joined_data$Population)) * 100
```

Fitting Linear Regression Model to the data

```{r}
model <- lm(`Premature age-adjusted mortality (Deaths)` ~ `Adult smoking (%)`+`Adult obesity (%)`+`Physical inactivity (%)`+Diabetics+`Poor mental health days`, data = joined_data)
display(model)
```

Creating tables to summarise key aspects of results obtained from linear regression model

```{r}
# Load required package
library(knitr)

# Create a data frame for the table
model_summary <- data.frame(
  coef.est = c(-0.49, 0.03, -0.01, 0.02, 0.27, 0.17),
  coef.se = c(0.04, 0.00, 0.00, 0.00, 0.01, 0.01),
  row.names = c("(Intercept)", "`Adult smoking (%)`", "`Adult obesity (%)`", "`Physical inactivity (%)`", "Diabetics (%)", "`Poor mental health days`")
)

# Create a data frame for the model metrics
model_metrics <- data.frame(
  n = 3062,
  k = 6,
  `residual sd` = 0.25,
  `R-Squared` = 0.66
)

# Print the summary table
kable(model_summary, caption = "Model Summary") %>%
  kable_styling()

# Print the model metrics
kable(model_metrics, caption = "Model Metrics") %>%
  kable_styling()

```

Saving the tables as csv files 

```{r}
# Write the tables to csv files
write.csv(model_summary, "model_summary.csv", row.names = TRUE)
write.csv(model_metrics, "model_metrics.csv", row.names = FALSE)
```

Creating scatter plot with linear regression line analysing premature age adjusted mortality with smoking

```{r}
plot1 <- ggplot(data = joined_data, aes(x = `Adult smoking (%)`, y = `Premature age-adjusted mortality (Deaths)`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color='brown') +
  labs(x = "Adult Smoking (%)", y = "Premature Age Mort.")
```

Creating scatter plot with linear regression line analysing premature age adjusted mortality with obesity

```{r}
plot2 <- ggplot(data = joined_data, aes(x = `Adult obesity (%)`, y = `Premature age-adjusted mortality (Deaths)`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color='brown') +
  labs(x = "Adult Obesity (%)", y = "Premature Age Mort.")
```

Creating scatter plot with linear regression line analysing premature age adjusted mortality with physical inactivity

```{r}
plot3 <- ggplot(data = joined_data, aes(x = `Physical inactivity (%)`, y = `Premature age-adjusted mortality (Deaths)`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color='brown') +
  labs(x = "Phycial Inactivity (%)", y = "Premature Age Mort.")
```

Creating scatter plot with linear regression line analysing premature age adjusted mortality with diabetes 

```{r}
plot4 <- ggplot(data = joined_data, aes(x = `Diabetics`, y = `Premature age-adjusted mortality (Deaths)`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color='brown') +
  labs(x = "Diabetics (%)", y = "Premature Age Mort.")
```

Creating scatter plot with linear regression line analysing premature age adjusted mortality with poor mental health days

```{r}
plot5 <- ggplot(data = joined_data, aes(x = `Poor mental health days`, y = `Premature age-adjusted mortality (Deaths)`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color='brown') +
  labs(x = "Poor mental health days", y = "Premature Age Mort.")
```

Combing all the 5 plots together

```{r}
library(patchwork)
combined_plots <- plot1 + plot2 + plot3 + plot4 + plot5 +
  plot_layout(nrow = 2)

# Display the combined plots
combined_plots
ggsave("combined_plots.png", combined_plots)
```


Calculating Pearson Correlation Coefficient for Premature Age Adjusted Mortality (P.A.A.M) and Smoking 

```{r}
cor(joined_data$`Premature age-adjusted mortality (Deaths)`, joined_data$`Adult smoking (%)`, use = "pairwise.complete.obs")
```

Calculating Pearson Correlation Coefficient for Premature Age Adjusted Mortality (P.A.A.M) and Obesity

```{r}
cor(joined_data$`Premature age-adjusted mortality (Deaths)`, joined_data$`Adult obesity (%)`, use = "pairwise.complete.obs")
```

Calculating Pearson Correlation Coefficient for Premature Age Adjusted Mortality (P.A.A.M) and Physical Inactivity

```{r}
cor(joined_data$`Premature age-adjusted mortality (Deaths)`, joined_data$`Physical inactivity (%)`, use = "pairwise.complete.obs")
```

Calculating Pearson Correlation Coefficient for Premature Age Adjusted Mortality (P.A.A.M) and Diabetes

```{r}
cor(joined_data$`Premature age-adjusted mortality (Deaths)`, joined_data$`Diabetics`, use = "pairwise.complete.obs")
```

Calculating Pearson Correlation Coefficient for Premature Age Adjusted Mortality (P.A.A.M) and Poor Mental Health Days

```{r}
cor(joined_data$`Premature age-adjusted mortality (Deaths)`, joined_data$`Poor mental health days`, use = "pairwise.complete.obs")
```

Creating a table displaying all the correlation coefficients 

```{r}
# Load required package
library(knitr)

# List of independent variables
independent_vars <- c("Adult smoking (%)", 
                      "Adult obesity (%)", 
                      "Physical inactivity (%)", 
                      "Diabetics", 
                      "Poor mental health days")

# Dependent variable
dependent_var <- "Premature age-adjusted mortality (Deaths)"

# Calculate correlations
correlations <- sapply(independent_vars, function(var) {
  cor(joined_data[[dependent_var]], joined_data[[var]], use = "pairwise.complete.obs")
})

# Create a data frame for the table
correlation_df <- data.frame(
  Independent_Variables = independent_vars,
  Correlation = correlations
)
correlation_df$Independent_Variables[4] <- "Diabetics (%)"
colnames(correlation_df) <- c("Predictors","Correlation with P.A.A.R.M")

# Display as a table in RMarkdown
kable(correlation_df, digits = 2)
```

Saving the above mentioned tables as a csv file

```{r}
write.csv(correlation_df, "correlation_df.csv", row.names = FALSE)
```

```{r}
saveRDS(correlation_df, "correlation_df.rds")
```



Reading the shapefile dataset to display the effect of above factors on P.A.A.M of counties throughout the nation

```{r}
sp <- st_read("/Users/bedi/Downloads/DATA5207_Project/Data/US shapefiles/cb_2016_us_county_20m.shp")

spf <- sp %>%
  dplyr::select(NAME, geometry)

spf <- spf %>%
  rename(County = NAME)
```

Joining the dataset with shapefiles data
```{r}
final_data <- left_join(joined_data, spf, by = "County")
```

Removing the duplicate county names from the final joined dataset

```{r}
# Check for duplicate county names in the spatial dataset
duplicate_county_spf <- spf[duplicated(spf$county), "County"]
print(duplicate_county_spf)

# Check for duplicate county names in the joined data
duplicate_county_joined <- joined_data[duplicated(joined_data$County), "County"]
print(duplicate_county_joined)

# Remove duplicate county names from the joined data (if necessary)
joined_data_unique <- joined_data[!duplicated(joined_data$County), ]

# Perform the join again using the unique joined data
joined_dat <- left_join(spf, joined_data_unique, by = c("County" = "County"))
```

Visualising effect of poor mental health days on counties in USA using mapview()

```{r}
map_poor <- mapview(joined_dat, zcol = "Poor mental health days", map.types = "OpenStreetMap")
```

```{r}
map_poor
```

Visualising effect of smoking on counties in USA using mapview()

```{r}
map_smo <- mapview(joined_dat, zcol = "Adult smoking (%)", map.types = "OpenStreetMap")
```

```{r}
map_smo
```

Visualising effect of obesity on counties in USA using mapview()

```{r}
map_ob <- mapview(joined_dat, zcol = "Adult obesity (%)", map.types = "OpenStreetMap")
```

```{r}
map_ob
```

Visualising effect of phsycial inactvity on counties in USA using mapview()

```{r}
map_phy <- mapview(joined_dat, zcol = "Physical inactivity (%)", map.types = "OpenStreetMap")
```

```{r}
map_phy
```

Visualising effect of diabetes on counties in USA using mapview()

```{r}
map_dia <- mapview(joined_dat, zcol = "Diabetics", map.types = "OpenStreetMap")
```

```{r}
map_dia
```

Visualising premature age adjusted mortality rates on counties in USA using mapview()

```{r}
map_age <- mapview(joined_dat, zcol = "Premature age-adjusted mortality (Deaths)", map.types = "OpenStreetMap")
```

```{r}
map_age
```

Creating maps combining premature age adjusted mortality of counties along with all the factors individually

```{r}
map_1 <- map_age + map_smo
map_1
```


```{r}
map_2 <- map_age + map_phy
map_2
```

```{r}
map_3 <- map_age + map_ob
map_3
```

```{r}
map_4 <- map_age + map_dia
map_4
```

```{r}
map_5 <- map_age + map_poor
map_5
```









